MATH 245 Homework 4

### 2024-03-25 PDEs/math245_hw4.Rmd
worked on pdes

Re-writing as $X''+\lambda X=0$, we will get the characteristic equation: $r^2+\beta^2=0$. Since our characteristic equation has complex roots $r= \pm i\beta$, our solutions take the form 
$$X(x)=A\sin{(\beta x)}+B\cos{(\beta x)}$$
Differentiating, we find that 
$$X'(x)=\beta A\cos{(\beta x)}-\beta B\sin{(\beta x)}$$
Now plugging in our initial condition $X'(0) = 0$, we get
$X'(0)=\beta A=0$. And since we are in a case where $\beta \neq 0$, this means $B=0$ so $X(x)=A\sin{(\beta x)}$. Now we use our boundary condition $X(l)=0$ to get $X(l)=A\sin{(\beta l)}=0$. If $A=0$, then $X(0)=0$ and this contradicts the definition of an eigenfunction. Therefore, we must have $\sin{(\beta l)}=0$, which can only occur if $\beta = \frac{(2n+1)\pi}{l}$. Therefore, this case gives us eigenvalues

$$\lambda_n= \left(\frac{(2n+1)\pi}{l}\right)^2 \quad n=1,2,3... $$
with eigenfunctions
$$X_n(x)=\sin{\left(\frac{(2n+1)\pi x}{l}\right)}$$

@@

$$X(x)=A\sin{(\beta \ln{x})}+B\cos{(\beta \ln{x})}$$
$$X(1)=A\sin{(0)}+B\cos{(0)}=0 \quad \longrightarrow \quad B=0$$
$$X(e)=A\sin{(\beta \ln{e})}=0$$
$$A\sin{(\beta)}=0$$
If $A=0$, then $X(0)=0$ and this contradicts the definition of an eigenfunction. Therefore, we must have $\sin{(\beta)}=0$, which can only occur if $\beta = (2n+1)\pi$. Therefore, this case gives us eigenvalues 
$$\lambda_n= (2n+1)^2 \pi^2 \quad n=1,2,3... $$
with eigenfunctions
$$X_n(x)=\sin{\left((2n+1)\pi \ln{x}\right)}$$

@@

Let's start with $a_0$:
\begin{align*}
a_0 &=\frac{1}{2\pi} \int_{-\pi}^{\pi}f(x)dx \\
    &=\frac{1}{2\pi}\left[ \int_{-\pi}^{-1} 1\;dx + \int_{-1}^{1} (1-\lvert x \rvert) dx+ \int_{1}^{\pi} 1\;dx  \right] \\
    &= \frac{1}{2\pi}\int_{-\pi}^{\pi}1\;dx - \int_{-1}^{1} \lvert x \rvert\, dx \\
    &= 1- 2\int_{0}^{1} x \; dx \\
    &= 1- 1=0
\end{align*}

@@

\begin{align*}
a_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\cos{\left(nx\right) }dx \\
    &= \frac{2}{\pi} \int_{0}^{\pi}x\cos{\left(nx\right) }dx \\
    &= \frac{2}{n \pi} x \sin{(nx)} \Bigr\rvert_0^{\pi} - \frac{2}{n \pi}\int_{0}^{\pi}\sin{\left( nx \right) } dx \\
    &= 0 + \frac{2}{n^2 \pi} \cos{(nx)}\Bigr\rvert_0^{\pi} \\
    &= \frac{2}{n^2 \pi} (-1)^{n+1}
\end{align*}

@@

\begin{align*}
a_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\cos{(nx)}dx \\
    &= \frac{1}{\pi} \int_{-\pi}^{\pi}x\cos{(nx)}+ \frac{1}{\pi} \int_{-\pi}^{\pi}x^2\cos{(nx)}dx 
\end{align*}

Applying integration by parts to the first integral,
$$\frac{1}{\pi} \int_{-\pi}^{\pi}x\cos{(nx)}=\frac{1}{n\pi}x\sin{(nx)} \Bigr\rvert_{-\pi}^{\pi}+\frac{1}{n\pi} \int_{-\pi}^{\pi}\sin{(nx)}$$
$$=\frac{1}{n\pi}x\sin{(nx)} \Bigr\rvert_{-\pi}^{\pi}+\frac{-1}{n^2\pi} \cos{(nx)}\Bigr\rvert_{-\pi}^{\pi}=0+0=0$$

We then use the integration by parts table method demonstrated in class to find that

$$\frac{1}{\pi} \int_{-\pi}^{\pi}x^2\cos{(nx)}= \frac{x^2}{n}\sin{(nx)}\Bigr\rvert_{-\pi}^{\pi} +  \frac{2x}{n^2}\cos{(nx)}\Bigr\rvert_{-\pi}^{\pi}+\frac{2}{n^3}\sin{(nx)}\Bigr\rvert_{-\pi}^{\pi}$$
The first and last terms (with $\sin{(nx)}$) vanish, while the middle term evaluates to $\frac{4 \cos{(n\pi)}}{n^2}$.

$$\frac{4 \cos{(n\pi)}}{n^2}=\frac{4}{n^2} (-1)^{n}$$
Thus, we add this to the result of first integral to find that
$$a_n= \frac{4}{n^2} (-1)^{n} $$
\begin{align*}
b_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\sin{(nx)}dx \\
    &= \frac{1}{\pi} \int_{-\pi}^{\pi}x\sin{(nx)}+ \frac{1}{\pi} \int_{-\pi}^{\pi}x^2\sin{(nx)}dx 
\end{align*}

@@

The Fourier sine series of a function $f(x)$ is $$f(x)= \sum_{n=1}^{\infty} A_n \sin{  \left(\frac{n\pi x}{l}\right) }$$
where 
$$A_n = \frac{2}{l} \int_{0}^{l}f(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx $$
$$A_n = \frac{2}{\pi} \int_{0}^{\pi}f(x)\sin{(nx)}dx $$
\begin{align*}
A_n &= \frac{2}{\pi}\left[ \int_{0}^{\pi/2}\sin{(nx)}dx + 2 \int_{\pi/2}^{\pi}\sin{(nx)}dx \right] \\
    &= \frac{2}{\pi}\left[\frac{-1}{n} \cos{(nx)}\Bigr\rvert_{0}^{\pi/2}- \frac{2}{n} \cos{(nx)}\Bigr\rvert_{\pi/2}^{\pi}\right] \\
    &= \frac{2}{n \pi}\left[\cos{\left(\frac{n\pi}{2}\right)}+1-2\cos{(n\pi)} \right]
\end{align*}

@@

$$a_n = \frac{1}{\pi} \int_{-\pi}^{\pi}x\cos{(nx)}dx$$

But in part (2b), we found that this evaluates to zero.

>>>>

Re-writing as $X''+\lambda X=0$, we will get the characteristic equation: $r^2+\beta^2=0$. Since our characteristic equation has complex roots $r= \pm i\beta$, our solutions take the form 
$$X(x)=A\sin{(\beta x)}+B\cos{(\beta x)}$$
Differentiating, we find that 
$$X'(x)=\beta A\cos{(\beta x)}-\beta B\sin{(\beta x)}$$
Now plugging in our initial condition $X'(0) = 0$, we get
$X'(0)=\beta A=0$. And since we are in a case where $\beta \neq 0$, this means $B=0$ so $X(x)=A\sin{(\beta x)}$. Now we use our boundary condition $X(l)=0$ to get $X(l)=A\sin{(\beta l)}=0$. If $A=0$, then $X(0)=0$ and this contradicts the definition of an eigenfunction. Therefore, we must have $\sin{(\beta l)}=0$, which can only occur if $\beta = \frac{n\pi}{l}$. Therefore, this case gives us eigenvalues

$$\lambda_n= \left(\frac{n\pi}{l}\right)^2 \quad n=1,2,3... $$
with eigenfunctions
$$X_n(x)=\sin{\left( \frac{n \pi x}{l}\right)}$$

@@

$$X(x)=A\sin{(\beta \ln{x})}+B\cos{(\beta \ln{x})}$$
$$X(1)=A\sin{(0)}+B\cos{(0)}=0 \quad \longrightarrow \quad B=0$$
$$X(e)=A\sin{(\beta \ln{e})}=0$$
$$A\sin{(\beta)}=0$$
If $A=0$, then $X(0)=0$ and this contradicts the definition of an eigenfunction. Therefore, we must have $\sin{(\beta)}=0$, which can only occur if $\beta = n\pi$. Therefore, this case gives us eigenvalues 
$$\lambda_n= n^2 \pi^2 \quad n=1,2,3... $$
with eigenfunctions
$$X_n(x)=\sin{\left(n\pi \ln{x}\right)}$$

@@

Here, $l=\pi$. Let's start with $a_0$:
\begin{align*}
a_0 &=\frac{1}{2\pi} \int_{-\pi}^{\pi}f(x)dx \\
    &=\frac{1}{2\pi}\left[ \int_{-\pi}^{-1} 1\;dx + \int_{-1}^{1} (1-\lvert x \rvert) dx+ \int_{1}^{\pi} 1\;dx  \right] \\
    &= \frac{1}{2\pi}\int_{-\pi}^{\pi}1\;dx - \int_{-1}^{1} \lvert x \rvert\, dx \\
    &= 1- 2\int_{0}^{1} x \; dx \\
    &= 1- 1=0
\end{align*}

@@

\begin{align*}
a_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\cos{\left(nx\right) }dx \\
    &= \frac{2}{\pi} \int_{0}^{\pi}x\cos{\left(nx\right) }dx \\
    &= \frac{2}{n \pi} x \sin{(nx)} \Bigr\rvert_0^{\pi} - \frac{2}{n \pi}\int_{0}^{\pi}\sin{\left( nx \right) } dx \\
    &= 0 + \frac{2}{n^2 \pi} \cos{(nx)}\Bigr\rvert_0^{\pi} \\
    &= \frac{2}{n^2 \pi} (-1)^{n}-\frac{2}{n^2 \pi} \\
    &= \frac{2}{n^2 \pi} (-1)^{n+1}
\end{align*}

@@

\begin{align*}
a_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\cos{(nx)}\;dx \\
    &= \frac{1}{\pi} \int_{-\pi}^{\pi}x\cos{(nx)}\;dx+ \frac{1}{\pi} \int_{-\pi}^{\pi}x^2\cos{(nx)}\;dx 
\end{align*}

Applying integration by parts to the first integral,
$$\frac{1}{\pi} \int_{-\pi}^{\pi}x\cos{(nx)}\;dx=\frac{1}{n\pi}x\sin{(nx)} \Bigr\rvert_{-\pi}^{\pi}-\frac{1}{n\pi} \int_{-\pi}^{\pi}\sin{(nx)}\;dx$$
$$=\frac{1}{n\pi}x\sin{(nx)} \Bigr\rvert_{-\pi}^{\pi}-\frac{-1}{n^2\pi} \cos{(nx)}\Bigr\rvert_{-\pi}^{\pi}=0-0=0$$

For the second integral, we can use the integration by parts table method demonstrated in class to find that

$$\frac{1}{\pi} \int_{-\pi}^{\pi}x^2\cos{(nx)}\;dx= \frac{x^2}{n\pi}\sin{(nx)}\Bigr\rvert_{-\pi}^{\pi} +  \frac{2x}{n^2\pi}\cos{(nx)}\Bigr\rvert_{-\pi}^{\pi}+
\frac{2}{n^3\pi}\sin{(nx)}\Bigr\rvert_{-\pi}^{\pi}$$
The first and last terms (with $\sin{(nx)}$) vanish. Evaluating the middle term, we get:

$$\frac{2\pi \cos{(n\pi)}}{n^2 \pi}-\frac{-2\pi \cos{(n\pi)}}{n^2 \pi}=\frac{4 \cos{(n\pi)}}{n^2}=\frac{4}{n^2} (-1)^{n}$$
Thus, we have found our coefficients $a_n$:
$$a_n= \frac{4}{n^2} (-1)^{n} $$
Now for $b_n$:
\begin{align*}
b_n &= \frac{1}{\pi} \int_{-\pi}^{\pi}f(x)\sin{(nx)}dx \\
    &= \frac{1}{\pi} \int_{-\pi}^{\pi}x\sin{(nx)}+ \frac{1}{\pi} \int_{-\pi}^{\pi}x^2\sin{(nx)}dx 
\end{align*}

@@

The Fourier sine series of a function $f(x)$ is $$f(x)= \sum_{n=1}^{\infty} A_n \sin{  \left(\frac{n\pi x}{l}\right) }$$
where 
$$A_n = \frac{2}{l} \int_{0}^{l}f(x)\sin{  \left(\frac{n\pi x}{l}\right) }dx $$
$$A_n = \frac{2}{\pi} \int_{0}^{\pi}f(x)\sin{(nx)}dx $$
\begin{align*}
A_n &= \frac{2}{\pi}\left[ \int_{0}^{\pi/2}\sin{(nx)}dx + 2 \int_{\pi/2}^{\pi}\sin{(nx)}dx \right] \\
    &= \frac{2}{\pi}\left[\frac{-1}{n} \cos{(nx)}\Bigr\rvert_{0}^{\pi/2}- \frac{2}{n} \cos{(nx)}\Bigr\rvert_{\pi/2}^{\pi}\right] \\
    &= \frac{2}{n \pi}\left[-\cos{\left(\frac{n\pi}{2}\right)}+1-2\cos{(n\pi)}+2\cos{\left(\frac{n\pi}{2}\right)} \right]\\
    &= \frac{2}{n \pi}\left[1+\cos{\left(\frac{n\pi}{2}\right)}-2\cos{(n\pi)} \right]
\end{align*}

@@

$$a_n = \frac{1}{\pi} \int_{-\pi}^{\pi}x^2\cos{(nx)}dx$$

But in part (2c), we found that this is equivalent to $\frac{4}{n^2} (-1)^{n}$, so
$$a_n= \frac{4}{n^2} (-1)^{n} $$
$$b_n = \frac{1}{\pi} \int_{-\pi}^{\pi}x^2\sin{(nx)}dx$$
Since $x^2$ is an even function, $b_n=0$.

$$f(x)=\frac{\pi^2}{3}+ 4 \sum_{n=1}^{\infty}  \frac{(-1)^{n}\cos{(nx) } }{n^2} $$
Now let $x=0$, such that $f(0)=0^2=0$ and $\cos{(nx)}=\cos{(0)}=1$.

$$0=\frac{\pi^2}{3}+ 4 \sum_{n=1}^{\infty}  \frac{(-1)^{n} }{n^2} $$
$$-\frac{\pi^2}{12}=\sum_{n=1}^{\infty}  \frac{(-1)^{n} }{n^2} $$
